Fractional Types

### 2013-02-11 frac.tex
_
relational semantics

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
%! [^ = \ulcorner
%! ^] = \urcorner
%! [v = \llcorner
%! v] = \lrcorner
%! [[ = \llbracket
%! ]] = \rrbracket
%! eta* = \eta
%! eps* = \epsilon
%! Union = \bigcup
%! in = \in
%! |-->* = \mapsto^{*}
%! |-->> = \mapsto_{\ggg}
%! |-->let = \mapsto_{let}
%! |--> = \mapsto
%! <--| = \mapsfrom
%! |- = \vdash
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! ::= = ::=
%! /= = \neq
%! vi = v_i
%! di = d_i
%! si = s_i
%! sj = s_j
%! F = \texttt{F}
%! T = \texttt{T}
%! forall = \forall
%! exists = \exists
%! empty = \emptyset
%! eta = \eta
%! where = \textbf{where}
%! epsilon = \varepsilon
%! least = \phi
%! loop+ = loop_{+}
%! loop* = loop_{\times}
%! CatC = {\mathcal C}
%! CatA = {\mathcal A}
%! gamma = \gamma
%! {[ = \{
%! ]} = \}
%! elem = \in
%! dagger = ^\dagger
%! alpha = \alpha
%! beta = \beta
%! rho = \rho
%! @@ = \mu
%! @ = \,@\,
%! langRev = \Pi
%! langRevT = \Pi^{o}
%! langRevEE = \Pi^{\eta\epsilon}_{+}
%! bullet = \bullet
%! * = \times
%! langRevT = \Pi^{o}
%! langRevTF = \Pi^{/}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  In previous work, we developed a \emph{first-order},
  information-preserving, and reversible programming language {{langRev}}
  founded on type isomorphisms. Being restricted to first-order types limits
  the expressiveness of the language: it is not possible, for example, to
  abstract common program fragments into a higher-level combinator. In this
  paper, we introduce a higher-order extension of {{langRev}} based on the
  novel concept of \emph{fractional types} {{1/b}}. Intuitively, a value of a
  fractional type {{1/v}} represents \emph{negative} information. A
  higher-order function is modeled by a pair {{(1/v1,v2)}} with {{1/v1}}
  representing the needed argument and {{v2}} representing the
  result. Fractional values are first-class: they can be freely propagated
  and transformed but must ultimately --- in a complete program --- be offset
  by the corresponding amount of positive information.
\end{abstract}

@@

In the computer science community, the theory and practice of type
isomorphisms is well-established. Originally, such type isomorphisms were
motivated by the pragmatic concern of searching large libraries of functions
by providing one of the many possible isomorphic types for the desired
function. More recently, type isomorphisms have taken a more central role as
\emph{the} fundamental computational mechanism from which more conventional,
i.e., irreversible computation, is derived~\cite{infeffects}. In more detail,
in our own previous work, we started with the notion of type isomorphism as
the ``right'' starting point for a theory of computation. The central
technical contribution was the development of a family of programming
languages, {{langRev}} with various superscripts, in which every computation
is an isomorphism. Other major results are that all such computations
preserve the information-theoretic entropy, that the language is logically
reversible, and that it has a model in symmetric monoidal categories (which
are the backbone of models for linear logic and quantum computation). We have
also invested a significant amount of time in using {{langRev}} as an actual
programming language and established several idioms for programming large
classes of interesting (recursive) programs in {{langRev}}.

A major open problem remains, however. This concerns a higher-order extension
of {{langRev}}. This extension is of fundamental importance in all the
originating research areas. In physics, it allows the process itself or the
observer to be treated as ``data'' that can be transformed by higher-level
processes or observed by meta-level observers. In mathematics and logic, it
allows the equivalence between different proofs of type $\texttt{Id}_A(a,b)$
to be itself expressed as an isomorphism as a higher type
$\texttt{Id}_{\texttt{Id}_A(a,b)}(p,q)$. Finally, in computer science,
higher-order types allow code to abstract over other code fragments as well
as the manipulation of code as data and data as code which is central to many
algorithms and application domains.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Background: {{langRev}} }
\label{sec:pi}

We review our language {{langRev}} providing the necessary background and
context for our higher-order extension.\footnote{The presentation in this
  section focuses on the simplest version of {{langRev}}. Other versions
  include the empty type, recursive types, and trace operators but these
  extensions are orthogonal to the higher-order extension emphasized in this
  paper.} The terms of {{langRev}} are not classical values and functions;
rather, the terms are isomorphism witnesses.  In other words, the terms of
{{langRev}} are proofs that certain ``shapes of values'' are isomorphic.
And, in classical Curry-Howard fashion, our operational semantics shows how
these proofs can be directly interpreted as actions on ordinary values which
effect this shape transformation. Of course, ``shapes of values'' are very
familiar already: they are usually called \emph{types}.  But usually one
designs a type system as a method of classifying terms, with the eventual
purpose to show that certain properties of well-typed terms hold, such as
safety.  Our approach is different: we start from a type system, and then
present a term language which naturally inhabits these types, along with an
appropriate operational semantics.

\paragraph*{Data.}
We view {{langRev}} as having two levels:  it has traditional values, given by
%subcode{bnf} include main
% values, v ::= () | left v | right v | (v, v)
\noindent and these are classified by ordinary types
%subcode{bnf} include main
% value types, b ::= 1 | b + b | b * b

@@

\paragraph*{Isomorphisms.} The terms of {{langRev}} are witnesses to type
isomorphisms.  They have (iso) types {{b <-> b}}.  
Specifically, they are witnesses to the following type isomorphisms:
%subcode{bnf} include main
%! columnStyle = r@{\hspace{-0.5pt}}c@{\hspace{-0.5pt}}l
%swap+ :&  b1 + b2 <-> b2 + b1 &: swap+
%assocl+ :&  b1 + (b2 + b3) <-> (b1 + b2) + b3 &: assocr+
%identl* :&  1 * b <-> b &: identr*
%swap* :&  b1 * b2 <-> b2 * b1 &: swap*
%assocl* :&  b1 * (b2 * b3) <-> (b1 * b2) * b3 &: assocr*
%dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor

\noindent Each line of the above table introduces a pair of dual constants%
\footnote{where {{swap*}} and {{swap+}} are self-dual} that witness the type
isomorphism in the middle.  These are the base (non-reducible) terms of the
second, principal level of {{langRev}}.

@@

\noindent
To summarize, the syntax of {{langRev}} is given as follows.

\begin{definition}{(Syntax of {{langRev}})}
\label{def:Pi}
%subcode{bnf} include main
% value types, b ::= 1 | b+b | b*b 
% values, v ::= () | left v | right v | (v,v) 
%
% iso.~types, t ::= b <-> b
% base iso ::= swap+ | assocl+ | assocr+ 
%     &|& unite | uniti | swap* | assocl* | assocr* 
%     &|& dist | factor 
% iso comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
% top level term, l ::= c v
\end{definition}

Mathematically, {{langRev}} can be interpreted in the category of finite sets
and bijections: each value type denotes a finite set of the corresponding size
and each combinator {{c : b1 <-> b2}} denotes a bijection between the sets
denoted by {{b1}} and {{b2}}. Operationally, the semantics of {{langRev}} is
given using two mutually recursive interpreters: one going forward and one
going backwards. The use of {{sym}} switches control from one evaluator to
the other. We will not present the operational semantics until the next
section, in the context of the extended language. For now, we state without
proof that the evaluation of well-typed combinators always terminates and
that {{langRev}} is logically reversible, i.e., that for all combinators 
{{c : b1 <-> b2}} and values {{v1 : b1}} and {{v2 : b2}} we have the forward
evaluation of {{c v1}} produces {{v2}} iff the backwards evaluation of 
{{c v2}} produces {{v1}}.

@@

Note that a small modification of {{langRev}} (the addition of an
empty type {{0}} with its usual laws) would exactly capture, at the type
level, the notion of \emph{semiring} (occasionally called a \emph{rig}) where
we replace equality by isomorphism.  This can also be also compared to
\emph{bimonoidal categories}.  As is usual with with type theories, the
coherence conditions (pentagonal and triangle identity) are not explicit,
but would correspond to certain identity types being trivial.  {{langRev}}
instead, at the type level, models a commutative ringoid where the
multiplicative structure forms a commutative monoid, but the additive structure
is just a commutative magma.  We chose not to include {{0}} as a technical
convenience for the next section.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Language: {{langRevTF}} }

The language {{langRev}} models isomorphisms of values rather well.  However,
purposefully, it has no notion of \emph{input} or \emph{output}; more
precisely, because of strict preservation of information, these two concepts
coincide.  This is the root of reversibility.

@@

\begin{definition}[Syntax of {{langRevTF}}]
\label{def:langRevT}
We extend Def.~\ref{def:Pi} with:
%subcode{bnf} include main
% value types, b ::= ... | 1/b
% values, v ::= ... | 1/v
%
% base iso ::= ... | eta*_b | eps*_b
\end{definition}

@@

In {{langRev}} presented in Sec.~\ref{sec:pi}, we have been able to maintain
the simple idea that a syntactic isomorphism denotes a \emph{bijection}. In
richer languages, e.g., in a Turing complete language that admits
non-termination, isomorphisms necessarily denote \emph{partial
  bijections}~\cite{rc2011,infeffects}. As we demonstrate in this section,
once we make the extension to higher-order functions, we must interpret the
syntactic isomorphisms as \emph{relations}. The intuitive idea, is that the
fission point combinator {{eta*_b : 1 <-> 1/b * b}} must be able to map the
element {{() : 1}} to any pair of the form {{(1/v,v)}} with
{{v:b}}. Similarly, the combinator {{eps*_b : 1/b * b <-> 1}} must be able to
accept pairs {{(1/v1,v2)}} in which the negative information {{1/v1}} does
not match the positive information {{v2}}. On such pairs, {{eps*_b}} is
undefined, i.e., it does not produce {{():1}}. In other words, some of the
combinators are not partial and some produce more than one result for a
particular input, and hence we move to the setting of general binary
relations to interpret our combinators.

This setting is not only consistent with logical reversibility but is
arguably the \emph{right} framework for considering expressive reversible
languages (including the interesting examples of quantum languages).  Indeed,
even though the categorical semantics of {{langRevTF}} is not the main focus
of this paper, we find it helpful to state that the category of finite sets
and relations in which we are going to interpret our language, is an example
of a \emph{biproduct dagger compact closed category} which are categories
introduced by Abramsky and Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878}
and by Selinger~\cite{Selinger:2007:DCC:1229185.1229207} to model quantum
computation.\footnote{The other common example of these categories is that of
  finite dimensional Hilbert spaces.} Two of the most relevant features of
these categories for our purposes are that they support an adjoint for each
morphism (to model reversibility) and that they support higher-order relation
(built from the categorical dagger which is modeled by our fractional
types). The latter is intuitively clear as any relation between $A$ and $B$
can be represented as a subset of {{A * B}}, i.e., as a value.

To interpret {{langRevTF}} in this category, we associate a finite set
{{ [[ b ]] }} with each type {{b}} and a relation {{ [[ c ]] }} over
{{ [[ b1 ]] }} and {{ [[ b2 ]] }} for each combinator {{ c : b1 <-> b2 }}.

@@

\begin{definition}[Relational semantics]
\label{def:relational-langRevTF}
Each combinator {{c : b1 <-> b2}} in {{langRevTF}} denotes a relation
as specified below.

@@

The language {{langRevTF}} is still logically reversible. This is intuitively
clear as the relation denoted by a combinator can be used in the reverse
direction by swapping inputs and outputs.

@@

@@

Mathematically speaking, recursion and iteration can be expressed using
categorical trace
operators~\cite{joyal1996traced,Hasegawa:1997:RCS:645893.671607}.  In a
language like {{langRev}} there are two natural families of trace operators
that can be defined, an additive family (explored in detail in our previous
work~\cite{rc2011}) and a multiplicative family (which is the focus of this
section).

@@

The above example should convince the reader that the language {{langRev}}
with multiplicative trace is expressive enough to model finite relational
programming. We will not prove this claim. We develop an even more
substantial example: a SAT solver.

@@

homotopy equivalence???

@@

>>>>

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include Pi
%! [^ = \ulcorner
%! ^] = \urcorner
%! [v = \llcorner
%! v] = \lrcorner
%! [[ = \llbracket
%! ]] = \rrbracket
%! eta* = \eta
%! eps* = \epsilon
%! Union = \bigcup
%! in = \in
%! |-->* = \mapsto^{*}
%! |-->> = \mapsto_{\ggg}
%! |-->let = \mapsto_{let}
%! |--> = \mapsto
%! <--| = \mapsfrom
%! |- = \vdash
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! ::= = ::=
%! /= = \neq
%! vi = v_i
%! di = d_i
%! si = s_i
%! sj = s_j
%! F = \texttt{F}
%! T = \texttt{T}
%! forall = \forall
%! exists = \exists
%! empty = \emptyset
%! eta = \eta
%! where = \textbf{where}
%! epsilon = \varepsilon
%! least = \phi
%! loop+ = loop_{+}
%! loop* = loop_{\times}
%! CatC = {\mathcal C}
%! CatA = {\mathcal A}
%! gamma = \gamma
%! {[ = \{
%! ]} = \}
%! elem = \in
%! dagger = ^\dagger
%! alpha = \alpha
%! beta = \beta
%! rho = \rho
%! @@ = \mu
%! @ = \,@\,
%! Pi = \Pi
%! PiT = \Pi^{o}
%! PiEE = \Pi^{\eta\epsilon}_{+}
%! bullet = \bullet
%! * = \times
%! PiT = \Pi^{o}
%! PiTF = \Pi^{/}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  In previous work, we developed a \emph{first-order},
  information-preserving, and reversible programming language {{Pi}}
  founded on type isomorphisms. Being restricted to first-order types limits
  the expressiveness of the language: it is not possible, for example, to
  abstract common program fragments into a higher-level combinator. In this
  paper, we introduce a higher-order extension of {{Pi}} based on the
  novel concept of \emph{fractional types} {{1/b}}. Intuitively, a value of a
  fractional type {{1/v}} represents \emph{negative} information. A
  higher-order function is modeled by a pair {{(1/v1,v2)}} with {{1/v1}}
  representing the needed argument and {{v2}} representing the
  result. Fractional values are first-class: they can be freely propagated
  and transformed but must ultimately --- in a complete program --- be offset
  by the corresponding amount of positive information.
\end{abstract}

@@

In the computer science community, the theory and practice of type
isomorphisms is well-established. Originally, such type isomorphisms were
motivated by the pragmatic concern of searching large libraries of functions
by providing one of the many possible isomorphic types for the desired
function. More recently, type isomorphisms have taken a more central role as
\emph{the} fundamental computational mechanism from which more conventional,
i.e., irreversible computation, is derived~\cite{infeffects}. In more detail,
in our own previous work, we started with the notion of type isomorphism as
the ``right'' starting point for a theory of computation. The central
technical contribution was the development of a family of programming
languages, {{Pi}} with various superscripts, in which every computation
is an isomorphism. Other major results are that all such computations
preserve the information-theoretic entropy, that the language is logically
reversible, and that it has a model in symmetric monoidal categories (which
are the backbone of models for linear logic and quantum computation). We have
also invested a significant amount of time in using {{Pi}} as an actual
programming language and established several idioms for programming large
classes of interesting (recursive) programs in {{Pi}}.

A major open problem remains, however. This concerns a higher-order extension
of {{Pi}}. This extension is of fundamental importance in all the
originating research areas. In physics, it allows the process itself or the
observer to be treated as ``data'' that can be transformed by higher-level
processes or observed by meta-level observers. In mathematics and logic, it
allows the equivalence between different proofs of type $\texttt{Id}_A(a,b)$
to be itself expressed as an isomorphism as a higher type
$\texttt{Id}_{\texttt{Id}_A(a,b)}(p,q)$. Finally, in computer science,
higher-order types allow code to abstract over other code fragments as well
as the manipulation of code as data and data as code which is central to many
algorithms and application domains.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Background: {{Pi}} }
\label{sec:pi}

We review our language {{Pi}} providing the necessary background and
context for our higher-order extension.\footnote{The presentation in this
  section focuses on the simplest version of {{Pi}}. Other versions
  include the empty type, recursive types, and trace operators but these
  extensions are orthogonal to the higher-order extension emphasized in this
  paper.} The terms of {{Pi}} are not classical values and functions;
rather, the terms are isomorphism witnesses.  In other words, the terms of
{{Pi}} are proofs that certain ``shapes of values'' are isomorphic.
And, in classical Curry-Howard fashion, our operational semantics shows how
these proofs can be directly interpreted as actions on ordinary values which
effect this shape transformation. Of course, ``shapes of values'' are very
familiar already: they are usually called \emph{types}.  But usually one
designs a type system as a method of classifying terms, with the eventual
purpose to show that certain properties of well-typed terms hold, such as
safety.  Our approach is different: we start from a type system, and then
present a term language which naturally inhabits these types, along with an
appropriate operational semantics.

\paragraph*{Data.}
We view {{Pi}} as having two levels:  it has traditional values, given by
%subcode{bnf} include main
% values, v ::= () | left v | right v | (v, v)
\noindent and these are classified by ordinary types
%subcode{bnf} include main
% value types, b ::= 1 | b + b | b * b

@@

\paragraph*{Isomorphisms.} The terms of {{Pi}} are witnesses to type
isomorphisms.  They have (iso) types {{b <-> b}}.  
Specifically, they are witnesses to the following type isomorphisms:
%subcode{bnf} include main
%! columnStyle = r@{\hspace{-0.5pt}}c@{\hspace{-0.5pt}}l
%swap+ :&  b1 + b2 <-> b2 + b1 &: swap+
%assocl+ :&  b1 + (b2 + b3) <-> (b1 + b2) + b3 &: assocr+
%identl* :&  1 * b <-> b &: identr*
%swap* :&  b1 * b2 <-> b2 * b1 &: swap*
%assocl* :&  b1 * (b2 * b3) <-> (b1 * b2) * b3 &: assocr*
%dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor

\noindent Each line of the above table introduces a pair of dual constants%
\footnote{where {{swap*}} and {{swap+}} are self-dual} that witness the type
isomorphism in the middle.  These are the base (non-reducible) terms of the
second, principal level of {{Pi}}.

@@

\noindent
To summarize, the syntax of {{Pi}} is given as follows.

\begin{definition}{(Syntax of {{Pi}})}
\label{def:Pi}
%subcode{bnf} include main
% value types, b ::= 1 | b+b | b*b 
% values, v ::= () | left v | right v | (v,v) 
%
% iso.~types, t ::= b <-> b
% base iso ::= swap+ | assocl+ | assocr+ 
%     &|& unite | uniti | swap* | assocl* | assocr* 
%     &|& dist | factor 
% iso comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
% top level term, l ::= c v
\end{definition}

Mathematically, {{Pi}} can be interpreted in the category of finite sets
and bijections: each value type denotes a finite set of the corresponding size
and each combinator {{c : b1 <-> b2}} denotes a bijection between the sets
denoted by {{b1}} and {{b2}}. Operationally, the semantics of {{Pi}} is
given using two mutually recursive interpreters: one going forward and one
going backwards. The use of {{sym}} switches control from one evaluator to
the other. We will not present the operational semantics until the next
section, in the context of the extended language. For now, we state without
proof that the evaluation of well-typed combinators always terminates and
that {{Pi}} is logically reversible, i.e., that for all combinators 
{{c : b1 <-> b2}} and values {{v1 : b1}} and {{v2 : b2}} we have the forward
evaluation of {{c v1}} produces {{v2}} iff the backwards evaluation of 
{{c v2}} produces {{v1}}.

@@

Note that a small modification of {{Pi}} (the addition of an
empty type {{0}} with its usual laws) would exactly capture, at the type
level, the notion of \emph{semiring} (occasionally called a \emph{rig}) where
we replace equality by isomorphism.  This can also be also compared to
\emph{bimonoidal categories}.  As is usual with with type theories, the
coherence conditions (pentagonal and triangle identity) are not explicit,
but would correspond to certain identity types being trivial.  {{Pi}}
instead, at the type level, models a commutative ringoid where the
multiplicative structure forms a commutative monoid, but the additive structure
is just a commutative magma.  We chose not to include {{0}} as a technical
convenience for the next section.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Language: {{PiTF}} }

The language {{Pi}} models isomorphisms of values rather well.  However,
purposefully, it has no notion of \emph{input} or \emph{output}; more
precisely, because of strict preservation of information, these two concepts
coincide.  This is the root of reversibility.

@@

\begin{definition}[Syntax of {{PiTF}}]
\label{def:PiT}
We extend Def.~\ref{def:Pi} with:
%subcode{bnf} include main
% value types, b ::= ... | 1/b
% values, v ::= ... | 1/v
%
% base iso ::= ... | eta*_b | eps*_b
\end{definition}

@@

In {{Pi}} presented in Sec.~\ref{sec:pi}, we have been able to maintain
the simple idea that a syntactic isomorphism denotes a \emph{bijection}. In
richer languages, e.g., in a Turing complete language that admits
non-termination, isomorphisms necessarily denote \emph{partial
  bijections}~\cite{rc2011,infeffects}. As we demonstrate in this section,
once we make the extension to higher-order functions, we must interpret the
syntactic isomorphisms as \emph{relations}. The intuitive idea, is that the
fission point combinator {{eta*_b : 1 <-> 1/b * b}} must be able to map the
element {{() : 1}} to any pair of the form {{(1/v,v)}} with
{{v:b}}. Similarly, the combinator {{eps*_b : 1/b * b <-> 1}} must be able to
accept pairs {{(1/v1,v2)}} in which the negative information {{1/v1}} does
not match the positive information {{v2}}. On such pairs, {{eps*_b}} is
undefined, i.e., it does not produce {{():1}}. In other words, some of the
combinators are not partial and some produce more than one result for a
particular input, and hence we move to the setting of general binary
relations to interpret our combinators.

This setting is not only consistent with logical reversibility but is
arguably the \emph{right} framework for considering expressive reversible
languages (including the interesting examples of quantum languages).  Indeed,
even though the categorical semantics of {{PiTF}} is not the main focus
of this paper, we find it helpful to state that the category of finite sets
and relations in which we are going to interpret our language, is an example
of a \emph{biproduct dagger compact closed category} which are categories
introduced by Abramsky and Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878}
and by Selinger~\cite{Selinger:2007:DCC:1229185.1229207} to model quantum
computation.\footnote{The other common example of these categories is that of
  finite dimensional Hilbert spaces.} Two of the most relevant features of
these categories for our purposes are that they support an adjoint for each
morphism (to model reversibility) and that they support higher-order relation
(built from the categorical dagger which is modeled by our fractional
types). The latter is intuitively clear as any relation between $A$ and $B$
can be represented as a subset of {{A * B}}, i.e., as a value.

To interpret {{PiTF}} in this category, we associate a finite set
{{ [[ b ]] }} with each type {{b}} and a relation {{ [[ c ]] }} over
{{ [[ b1 ]] }} and {{ [[ b2 ]] }} for each combinator {{ c : b1 <-> b2 }}.

@@

\begin{definition}[Relational semantics]
\label{def:relational-PiTF}
Each combinator {{c : b1 <-> b2}} in {{PiTF}} denotes a relation
as specified below.

@@

The language {{PiTF}} is still logically reversible. This is intuitively
clear as the relation denoted by a combinator can be used in the reverse
direction by swapping inputs and outputs.

\begin{verbatim}
In the rules above it is unclear as to which v is chosen in the rule
 for eta. It is unclear that these rules are logically reversible.
\end{verbatim}

@@

\subsection{Relational Semantics}

Some thoughts on a relational semantics. We define {{ a <-> b }}
$\equiv$ {{a R b}}

\noindent
We now have primitives:
%subcode{opsem} include main
% swap* '= {[ ((v1, v2), (v2, v1)) ~|~ v1 in b1, v2 in b2 ]}
% swap+ '= {[ ((left v1), (right v1)) and ((right v2), (left v2)) ~|~ v1 in b1, v2 in b2]}
% assocl* '= {[ ((v1, (v2, v3)), ((v1, v2), v3)) ~|~ v1 in b1, v2 in b2, v3 in b3]}
etc..

\noindent
closure combinators:
%subcode{opsem} include main
% sym c '= {[(v2, v1) ~|~ (v1, v2) in c ]}
% c1 (;) c2 '= {[(v1, v3) ~|~ v2 in b2, (v1, v2) in c1, (v2, v3) in c2 ]}
% c1 (*) c2 '= {[((v1, v2), (v3, v4)) ~|~ (v1, v3) in c1, (v2, v4) in c2 ]}
etc..

\noindent
Now we have 
%subcode{opsem} include main
% eta* '= {[ ((), (v, 1/v)) ~|~ v in b ]}
% eps* '= {[ ((v, 1/v), ()) ~|~ v in b ]}

Now {{PiTF}} is logically reversible in that for all {{c:b1<->b2}} we
have {{c^{dagger}: b2 <-> b1}} such that 
%
{{c^{dagger} = {[(v2, v1)~|~(v1,v2) in c ]} }}. This is
straightforward from the definitions above.

We can define evaluation as follows: 
%
{{ c v = {[ v' ~|~ (v, v') in c ]} }}.

However this is unpleasant since it is unclear how to evaluate if
{{c~v}} and {{c^{dagger} v'}} are dual by evaluation. This ties into a
deeper issue of interpreting relations as programs.

Q) If {{|b1|=|b2|}}, can we define all relations between {{b1}} and
{{b2}}? If yes, then we have a model for {{PiTF}} where the objects
are sets {{b}} and the morphisms are the relations between objects of
equal size.

@@

Mathematically speaking, recursion and iteration can be expressed using
categorical trace
operators~\cite{joyal1996traced,Hasegawa:1997:RCS:645893.671607}.  In a
language like {{Pi}} there are two natural families of trace operators
that can be defined, an additive family (explored in detail in our previous
work~\cite{rc2011}) and a multiplicative family (which is the focus of this
section).

@@

The above example should convince the reader that the language {{Pi}}
with multiplicative trace is expressive enough to model finite relational
programming. We will not prove this claim. We develop an even more
substantial example: a SAT solver.

@@

\begin{verbatim}
homotopy equivalence???

@@

\end{verbatim}
